---
title: "Lecture 2 Code"
author: "Alex Hoagland"
date: "September 21, 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r header}
################################################################################
# I like to include several additional notes in the header of my files here:
#
# Last modified: 9/20/2024
#
### PURPOSE:
  # Lecture 2 code and output file
# 
### NOTES:
  # - uses the Tidyverse package and Dplyr
################################################################################

### 0. Load the packages we will need for this file ####
library(tidyverse) # load the installed package for each new session of R
# library(here) # for file organization
library(broom) # helps for storing regression output

set.seed(122333) # Setting the seed helps to make random number generators give us the same numbers across machines
################################################################################
```

## Regression

### Preliminaries

Let's first set up our data in the case where we **know** our data generating process (DGP). Pretend that we know that

$$ y = 0+5.5x+\varepsilon $$

for some random variation $\varepsilon$.

```{r data-setup}
tb <- tibble(
  x = rnorm(10000),
  u = rnorm(10000),
  y = 5.5*x + 12*u
) # Here's our true model, with some randomness baked in

# What is our expectation of x? y? u? What about variance? 
mean(tb$x)
mean(tb$y)
mean(tb$u)

var(tb$x)
var(tb$y)
var(tb$u)
```

In this case, what happens when we run a regression of $y$ on $x$?

```{r simple-reg}
tb %>% lm(y ~ x, .) 
```

### Visualizing Regressions

There's **a lot** we can do in R with regressions. Let's try some!

```{r model-viz}
# Looking at your model output
reg_tb <- lm(y ~ x, data=tb) # Store regression as an object 

summary(reg_tb) # gives you a full summary
regdata <- broom::tidy(reg_tb, conf.int=TRUE) # Stores regression output in a data frame
regouts <- broom::glance(reg_tb) # Stores other regression features in a data frame
```

Rather than just seeing what the regression returns (e.g., in a table), we can also look at its performance in the data (e.g., with more data visualizations)

```{r reg-performance}
tb <- tb %>% 
  mutate(
    yhat1 = predict(lm(y ~ x, .)),
    yhat2 = -0.14 + 5.73*x, 
    uhat1 = residuals(lm(y ~ x, .)),
    uhat2 = y - yhat2
  ) # How close are our predictions? 

summary(tb[-1:-3])
```

The most common visualization, especially for a one-dimensional regression, is just to plot the line against the data. Look closely at this command and let's walk through the syntax as we develop the plot.

```{r reg-viz-2}
tb %>% 
  lm(y ~ x, .) %>% 
  ggplot(aes(x=x, y=y)) + 
  ggtitle("OLS Regression Line") +
  geom_point(size = 0.05, color = "black", alpha = 0.5) +
  geom_smooth(method = lm, color = "black") +
  annotate("text", x = -1.5, y = 30, color = "red", 
           label = paste("Intercept = ", -0.14)) +
  annotate("text", x = 1.5, y = -30, color = "blue", 
           label = paste("Slope =", 5.73)) + 
  theme_classic() + labs(x="X", y="Y")
```

### Measures of Regression performance

Let's calculate the R-squared by hand first:

```{r r2-manual}
SST <- tb %>% mutate(ybar=mean(y), sst=(y-ybar)^2) %>% select(sst) %>% summarise(sum(sst))
SSE <- tb %>% mutate(ybar=mean(y), sse=(yhat1-ybar)^2) %>% select(sse) %>% summarise(sum(sse))
SSR <- tb %>% mutate(ssr=uhat1^2) %>% select(ssr) %>% summarise(sum(ssr))

R2 <- SSE/SST
print(paste0("The R-squared is ", round(as.numeric(R2)*100,digits=2), "%", sep=""))
```

You can also retrieve it straight from the lm command (this is almost always true for whatever regression object you might want!)

```{r r2}
summary(reg_tb) # Full summary
round(summary(reg_tb)$r.squared,digits=4) # Specifically, the r-squared
```

### Unbiasedness of our regressions

We've already shown that our model does reasonably well matching the truth, but can we show that in general? Let's try another simulation.

First, let's start with a population model (which we assume we know, in order to show that our regression can approach it):

$$ y = 3 + 2x + \varepsilon.$$ We will assume that $x$ and $\varepsilon$ are normally distributed, with $x \sim \mathcal{N}(0,9)$ and $\varepsilon \sim \mathcal{N}(0,36)$. We assume all the standard regression assumptions are met, so that $x$ and $\varepsilon$ are independent.

Let's use this DGP to create a random sample of data ($N$=10,000) and run a regression. We will repeat this procedure 1,000 times, storing the regression coefficients each time.

```{r unbiasedness}
lm <- lapply( # lapply is a useful command (see ?lapply to access documentation)
  1:10000,
  function(x) tibble(
    x = 9*rnorm(10000), # Distribution for x (rnorm(N) creates N random draws from a N(0,1) distribution)
    u = 36*rnorm(10000), # Distribution for e
    y = 3 + 2*x + u
  ) %>% 
    lm(y ~ x, .)
) # This runs 10,000 regressions with new data every time
# lapply is useful here since we want to resample data *and* do a regression at the same time. 
```

What are the results of this simulation? Let's make a table and a graph.

```{r unbiasedness-results}
as_tibble(t(sapply(lm, coef))) %>%
  summary(x) # Summarize the regression coefficients (tabular)

as_tibble(t(sapply(lm, coef))) %>% 
  ggplot()+
  geom_histogram(aes(x), binwidth = 0.01,fill="gray",color='black') + 
  theme_minimal() + 
  labs(x="Estimated Slope Coefficient", y="Count",title="Distribution of Slope Coefficients") +
  geom_vline(xintercept=2,color='red')
```

### Omitted variable bias

We just showed OLS is unbiased *when the regression assumptions hold*. But what happens when they **don't** -- specifically, when we leave a variable out of the model that belongs in it?

Imagine a new Type 2 diabetes drug. Sicker patients (higher `severity`) are more likely to be put on the drug **and** to have worse blood-sugar control (higher `A1c`). Suppose the **true** DGP is

$$ A1c = 8 - 0.8 \times Drug + 1.2 \times Severity + \varepsilon. $$

So the drug truly *lowers* A1c by 0.8 points. The problem is that `severity` is a confounder: it drives both who gets the drug and the outcome. Let's simulate this.

```{r ovb-dgp}
set.seed(5744)
n <- 2000
ovb <- tibble(
  severity = rnorm(n),                             # latent sickness (unobserved by naive analyst)
  drug     = rbinom(n, 1, plogis(1.3 * severity)), # sicker patients more likely to be treated
  A1c      = 8 - 0.8*drug + 1.2*severity + rnorm(n, 0, 0.5) # TRUE drug effect is -0.8
)
```

If we regress `A1c` on `drug` alone -- omitting `severity` -- what do we get?

```{r ovb-naive}
naive <- lm(A1c ~ drug, data = ovb)
broom::tidy(naive) # the drug coefficient is POSITIVE: the drug looks harmful!
```

Now put the omitted confounder back in:

```{r ovb-adjusted}
adjusted <- lm(A1c ~ drug + severity, data = ovb)
broom::tidy(adjusted) # the drug coefficient flips to about -0.8: the true effect
```

The estimate didn't just shrink -- it **changed sign**. And we can predict the bias exactly. The omitted-variable-bias formula says

$$ \text{plim}(\hat\beta_{drug}^{\,short}) = \beta_{drug} + \beta_{severity}\times\delta, $$

where $\delta$ is the slope from regressing the *omitted* variable on the *included* one:

```{r ovb-decomposition}
delta <- coef(lm(severity ~ drug, data = ovb))["drug"] # how severity differs by treatment
b_sev <- coef(adjusted)["severity"]                    # effect of severity on A1c

# The predicted bias should exactly equal the (naive - adjusted) drug coefficient:
c(predicted_bias = as.numeric(b_sev * delta),
  actual_gap     = as.numeric(coef(naive)["drug"] - coef(adjusted)["drug"]))
```

The two numbers match: the *entire* gap between the naive and adjusted estimates is accounted for by the omitted variable. This is exactly why we need **multivariate regression** -- our next topic.

## Inference in regression

### How to obtain standard errors
Just like a lot of other things, standard errors are stored in the lm object! 

```{r ses}
myout <- summary(reg_tb)
myout$coefficients[,2]
```

### Hypothesis testing

Let's start with a population model where: 
$$ y = -1 - 0.5x + \varepsilon, \varepsilon \sim \mathcal{N}(0,1).$$

We will use this DGP to create a small data set ($N=200$) and run a simple regression. Then, we will save the coefficient $\beta_1$ and its standard error. 
```{r hyp-test}
set.seed(0326)
nulldata <- tibble(x=rnorm(200),
                 epsilon=rnorm(200),
                 y=-1-.5*x+epsilon)
m1 <- lm(y ~ x,data=nulldata)
out <- summary(m1) # I get a first estimate of beta = , and a SE = 
myse <- out$coefficients[2,2] # store the SE
```

Our test, formally, is of the null hypothesis: 

$$ \mathcal{H}_0: \beta_1 = 0. $$
Let's plot our regression estimate against the sampling distribution of the null hypothesis
```{r hyp-test-2} 
myalf <- .05 # State the rejection probability

# Plot my estimate against the sampling distribution of the null hypothesis 
graphdata <- tibble(x=rnorm(1e4,mean=0,sd=myse), # normal distribution
                    xrej = ifelse(x >= qnorm(1-(myalf)/2, mean=0, sd=myse),x,NA), # x's that fall in the rejection region based on alpha
                    xrej2 = ifelse(-x >= qnorm(1-(myalf)/2, mean=0, sd=myse),x,NA), # x's that fall in the rejection region based on alpha
                    y=dnorm(x,mean=0,sd=myse)
)

 ggplot(data=graphdata) + 
  geom_density(aes(x)) + 
  geom_area(data=subset(graphdata, !is.na(xrej)),aes(x=x,y=y),fill="gray") + 
   geom_area(data=subset(graphdata, !is.na(xrej2)),aes(x=x,y=y),fill="gray") + 
  geom_vline(xintercept=-0.5,linetype='dashed',color='red') + 
  theme_minimal() + 
  labs(x="Estimated Beta",y="Density")
```

So what do we conclude from the test? 

#### One-sided testing

What happens if I perform a one-sided test? This changes the rejection region (is that disingenious?)
```{r hyp-test-3} 
graphdata <- tibble(x=rnorm(1e5,mean=0,sd=myse), # normal distribution
                     xrej = ifelse(x >= qnorm(1-(myalf), mean=0, sd=myse),x,NA), 
                        # x's that fall in the rejection region based on alpha
                     y=dnorm(x,mean=0,sd=myse)
 )
 
 ggplot(data=graphdata) + 
   geom_density(aes(x)) + 
   geom_area(data=subset(graphdata, !is.na(xrej)),aes(x=x,y=y),fill="gray") + 
   geom_vline(xintercept=-0.5,linetype='dashed',color='red') + 
   theme_minimal() + 
   labs(x="Estimated Beta",y="Density")
```

#### Statistical power

In general, adding more data adds to our statistical power -- the probability that we will uncover an effect that is truly in the DGP. Let's increase our sample size to 20,000 and see how this affects our standard error (and hence, our test statistic)

```{r hyp-test-4}
 nulldata <- tibble(x=rnorm(20000),
                  epsilon=rnorm(20000),
                  y=-1-.5*x+epsilon)
 m1 <- lm(y ~ x,data=nulldata)
 out <- summary(m1) # 
 out
 myse <- out$coefficients[2,2] # store the SE
 
 graphdata <- tibble(x=rnorm(1e5,mean=0,sd=myse), # normal distribution
                     xrej = ifelse(x >= qnorm(1-(myalf)/2, mean=0, sd=myse),x,NA), 
                        # x's that fall in the rejection region based on alpha
                     xrej2 = ifelse(-x >= qnorm(1-(myalf)/2, mean=0, sd=myse),x,NA), 
                        # x's that fall in the rejection region based on alpha
                     y=dnorm(x,mean=0,sd=myse)
 )
 
 ggplot(data=graphdata) + 
   geom_density(aes(x)) + 
   geom_area(data=subset(graphdata, !is.na(xrej)),aes(x=x,y=y),fill="gray") + 
   geom_area(data=subset(graphdata, !is.na(xrej2)),aes(x=x,y=y),fill="gray") + 
   geom_vline(xintercept=-0.5,linetype='dashed',color='red') + 
   theme_minimal() + 
   labs(x="Estimated Beta",y="Density")
 
# Calculate p-value of our estimated beta
1-pnorm(out$coefficients[2,1],mean=0,sd=out$coefficients[2,2])

# Calculate the t-statistic of our estimated beta
out$coefficients[2,1]/out$coefficients[2,2] # Here, critical value is 1.96
```
