---
title: "Lecture 5 Code"
author: "Alex Hoagland"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r header}
################################################################################
# I like to include several additional notes in the header of my files here:
#
# Last modified:  10/10/2024
#
### PURPOSE:
  # Lecture 5 code and output file
# 
### NOTES:
  # - uses the Tidyverse package and Dplyr
################################################################################


### 0. Load the packages we will need for this file ####
library(tidyverse) # load the installed package for each new session of R
library(ipumsr) # used to read in data from IPUMS
library(here) # we'll be talking about R projects and saving output here

set.seed(03262020) # Setting the seed helps to make random number generators give us the same numbers across machines
################################################################################
```

# Lecture 5: Data Cleaning and Research Design

We will use a sample dataset from the MEPS (Medical Expenditure Panel Survey) dataset, stored on IPUMS (see the slides for the links). Our question of interest will be: how does a health shock affect healthcare utilization for oneself and their spouse? 

First, we load the data

## Load data
When we make an extract request from IPUMS, they give us this code to pull our data into R: 
```{r load-data}
# NOTE: To load data, you must download both the extract's data and the DDI
# and also set the working directory to the folder with these files (or change the path below).

if (!require("ipumsr")) stop("Reading IPUMS data into R requires the ipumsr package. It can be installed using the following command: install.packages('ipumsr')")

ddi <- read_ipums_ddi("meps_00006.xml")
data <- read_ipums_micro(ddi)

data <- data %>% mutate(STATE=sample.int(50,size=nrow(data),replace=TRUE)) # Random state identifier for teaching purposes later on
```

## Data cleaning
What does data cleaning look like? It can look like several things: 
    - Removing missing data
    - Recoding variables
    - Creating new variables
    - Summarizing variables
    - Creating descriptive figures
    
Don't underestimate the value of descriptive evidence! 

In our project, we are looking at spending as our outcome variables, and health shocks as our index events. Let's look at these more carefully: 

```{r data-cleaning}
# First, outcome variables: total expenditures and total visits (you can play around with these)
# For now, we will just look at EXPTOT
data$EXPTOT # Note that IPUMS includes summaries of the variables here, really handy
data$EXPTOT %>% summary()  # What do we see here? 
sum(is.na(data$EXPTOT)) # check for missing values
hist(data$EXPTOT) # Quiz: how would we clean this up for a ggplot figure?
  # What does this tell us about a research design? 

# Let's topcode and inflation-adjust this variable 
  # why would we do these? 
data$EXPTOT <- data$EXPTOT * 1.23 # inflation adjust from 2019 to 2024 USD (https://www.in2013dollars.com/us/inflation/2019?amount=1)
data$EXPTOT <- ifelse(data$EXPTOT > 250000, 250000, data$EXPTOT) # topcode
hist(data$EXPTOT)
data <- data %>% mutate(logEXPTOT = log(EXPTOT)) # log-transform for analysis
data$EXPTOT %>% summary() 

# Now cause we're curious -- what fraction of individuals have 0 spending (and does it make sense?)
data %>% filter(EXPTOT == 0) %>% nrow() / nrow(data) # Does this seem reasonable? 

# Next, we will look at the health shock variable: CANCEREV
data$CANCEREV %>% summary() # What do we see here? This isn't helpful!
data$CANCEREV # this gives us the labels
data$CANCEREV %>% table() # This is better -- about 9% have ever had cancer

# Let's create our shock variable
data <- data %>% mutate(healthshock = as.numeric(CANCEREV == 2))
hist(data$healthshock)

# What about more specific shocks?
data <- data %>% mutate(healthshock_breast = as.numeric(CNBRES == 2), 
                        healthshock_colon = as.numeric(CNCOLN == 2), 
                        healthshock_lung = as.numeric(CNLUNG == 2), 
                        healthshock_pros = as.numeric(CNPROS == 2))
data %>% select(starts_with("healthshock")) %>% summary() # What does this tell us about the questions we can answer?

# Now briefly, what about some raw descriptive plots (across the whole sample)
# Note: this isn't quite descriptive evidence yet -- we would need to recreate these across some analytic sample
cor(data$healthshock, data$EXPTOT) # There is a positive correlation -- is it informative enough? 
ggplot(data, aes(x = factor(healthshock), y = EXPTOT)) + geom_boxplot() + theme_minimal() # What do we see here? 
ggplot(data, aes(x = factor(healthshock), y = log(EXPTOT))) + geom_boxplot() + theme_minimal() # Is this better? Why/why not? 

# Finally, do this across all types of cancers/health shock variables
# Note: this is a little trickier given the way we've constructed our data
data %>% mutate(healthshock_none = ifelse(healthshock==1,0,1)) %>% 
  select(MEPSID,starts_with("healthshock"),EXPTOT) %>% 
  pivot_longer(cols=starts_with("healthshock")) %>% # summary()
  filter(value == 1) %>% # keep spending in each category only if it is equal to 1
  ggplot(aes(x = factor(name), y = log(EXPTOT), fill = name)) + 
    geom_boxplot() + theme_minimal() # Note: I modified some copilot code to get this. 

  # Play around with this: how can you get it nicer? 
```

## Data Merging
Often, your project won't require just one dataset -- you may want to bring in publicly-available information or link multiple datasets in your work. How do we do this? 

``` {r data-merge}
# Let's say we want to merge in some publicly available data on cancer rates by state

# First, we need to get the data
cancer_rates <- read.csv("https://data.cdc.gov/api/views/u4d7-xz8k/rows.csv?accessType=DOWNLOAD") # Note: copilot just gave this to me! I didn't even have to google it! 
  # But is it perfect? No, I'm sure there's better 2019 cancer data available. 

cancer_rates <- cancer_rates %>% filter(Year == 2017) # we only need one year
levels(factor(cancer_rates$Cause.Name)) # we have more than cancer deaths here
cancer_rates <- cancer_rates %>% filter(Cause.Name == "Cancer")

# How do we get this into a format that will merge with our original data? 
library(cdlTools) # this has a "fips" command that let us convert to states
cancer_rates <- cancer_rates %>% mutate(STATE = fips(cancer_rates$State,to="FIPS"))
tomerge <- cancer_rates %>% select(STATE,Age.adjusted.Death.Rate) # this is all we want to merge in 
data <- data %>% left_join(tomerge, by = "STATE") 
  # Note: we're using a left join here -- why? 
data %>% select(Age.adjusted.Death.Rate) %>% summary() # What do we see here? Note there are some missing values! Why? data %>% filter(is.na(Age.adjusted.Death.Rate)) %>% select(STATE) %>% table()

# Now let's check if spending for cancer is higher in states with higher death rates using a scatterplot
ggplot(data %>% filter(healthshock == 1), aes(x = Age.adjusted.Death.Rate, y = log(EXPTOT))) + geom_point() + theme_minimal() # What do we see here? Does it make sense? 
``` 

## Constructing an Analytic Dataset
Now, we're ready to do some analysis! We have two cohorts we want to look at -- those affected by healthshocks and their spouses. Let's construct two datasets for these two cohorts.  

Remember, while we're doing this, we should *never* save over the original data. 

``` {r analytic-dataset}
# First, let's make a dataset of individuals who have had a health shock and those who haven't. Since we care about within-family spillovers, the "control" group should exclude those with spouses who have had a health shock as well. 

# What other things should we include? Let's (1) keep only households with 2 adults and (2) limit to adults aged 18+ for now. Why? 
  # Note: do we have a real control group here? Why/why not?
cohort1_treated <- data %>% filter(AGE >= 18) %>% group_by(DUID,FAMIDYR) %>% 
  mutate(famsize = n()) %>% filter(famsize >= 2) %>% 
  filter(healthshock == 1) %>% mutate(treated = 1) # individuals who have had a health shock
cohort1_control <- data %>% filter(AGE >= 18) %>% group_by(DUID,FAMIDYR) %>% 
  mutate(famsize = n()) %>% filter(famsize >= 2) %>% 
  mutate(todrop = max(healthshock)) %>%
  filter(todrop == 0) %>% ungroup() %>% select(-todrop) %>% mutate(treated = 0)
  # individuals who haven't had a health shock *in their household*
cohort1 <- rbind(cohort1_treated,cohort1_control)

# Now the second cohort can be the same control group, but now the "treated" group is *spouses* of those with cancer
cohort2_treated <- data %>% filter(AGE >= 18) %>% group_by(DUID,FAMIDYR) %>% 
  mutate(famsize = n()) %>% filter(famsize >= 2) %>% 
  mutate(anyshock = max(healthshock), 
         healthshock_breast = max(healthshock_breast),
         healthshock_colon = max(healthshock_colon),
         healthshock_lung = max(healthshock_lung),
         healthshock_pros = max(healthshock_pros)) %>% # note we also want flags for the *type of cancer* affecting the household
  filter(anyshock == 1 & healthshock == 0) %>% 
  mutate(treated = 1) # individuals whose spouses have had a health shock
cohort2 <- rbind(cohort2_treated,cohort1_control)

rm(cohort1_treated,cohort2_treated,cohort1_control)
```

## Analysis and Reporting Results
We'll spend a lot of time in this section in class as our methodologies evolve. For now, let's just do some more descriptive work across our cohorts. 

``` {r analysis}
# Let's start by looking at the distribution of spending across our cohorts
cohort1$treated <- factor(cohort1$treated,levels=c(0,1),labels=c("No Health Shock","Health Shock"))
cohort1 %>% ggplot(aes(x = log(EXPTOT), fill = factor(treated))) + 
  geom_density(alpha = 0.5) + theme_minimal() + 
  labs(title = "Distribution of Spending by Health Shock Status", 
       x="Log(Total Annual Spending)",
       y="",
       fill = "Own Health Shock Status")

# Can we make this more interpretable? What would a good paper figure need?
mean1 <- cohort1 %>% ungroup() %>% filter(treated=="Health Shock") %>% 
  summarize(mean = median(EXPTOT,na.rm=T)) %>% as.numeric()
mean0 <- cohort1 %>% ungroup() %>% filter(treated=="No Health Shock") %>% 
  summarize(mean = median(EXPTOT,na.rm=T)) %>% as.numeric()
logmean1 <- log(mean1)
logmean0 <- log(mean0)
fig1 <- cohort1 %>% ggplot(aes(x = log(EXPTOT), fill = factor(treated))) + 
  geom_density(alpha = 0.5) + theme_classic() + 
  geom_vline(xintercept = logmean1, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = logmean0, linetype = "dashed", color = "red") +
  annotate("text", x = logmean1+1.8, y = 0.3, label = paste("Median: $",round(mean1,0))) +
  annotate("text", x = logmean0-1.8, y = 0.3, label = paste("Median: $",round(mean0,0))) +
  labs(title = "Distribution of Spending by Health Shock Status", 
       x="Log(Total Annual Spending)",
       y="",
       fill = "Own Health Shock Status")
ggsave(fig1, file=here("fig1.png"), width=8, height=6, units="in", dpi=300)

# Now what about doing the same thing for spouses?
cohort2$treated <- factor(cohort2$treated,levels=c(0,1),labels=c("No Health Shock","Spouse Health Shock"))
mean1 <- cohort2 %>% ungroup() %>% filter(treated=="Spouse Health Shock") %>% 
  summarize(mean = median(EXPTOT,na.rm=T)) %>% as.numeric()
mean0 <- cohort2 %>% ungroup() %>% filter(treated=="No Health Shock") %>% 
  summarize(mean = median(EXPTOT,na.rm=T)) %>% as.numeric()
logmean1 <- log(mean1)
logmean0 <- log(mean0)
fig2 <- cohort2 %>% ggplot(aes(x = log(EXPTOT), fill = factor(treated))) + 
  geom_density(alpha = 0.5) + theme_classic() + 
  geom_vline(xintercept = logmean1, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = logmean0, linetype = "dashed", color = "red") +
  annotate("text", x = logmean1+1.8, y = 0.3, label = paste("Median: $",round(mean1,0))) +
  annotate("text", x = logmean0-1.8, y = 0.3, label = paste("Median: $",round(mean0,0))) +
  labs(title = "Distribution of Spending by Spousal Health Shock Status", 
       x="Log(Total Annual Spending)",
       y="",
       fill = "Spousal Health Shock Status")
ggsave(fig2, file=here("fig2.png"), width=8, height=6, units="in", dpi=300)

# Finally, we can ask the question: do spouses have different health spending based on cancer types? 
fig3 <- cohort1 %>% ungroup() %>% mutate(healthshock_none = ifelse(treated==0,1,0),
                                 healthshock_any = ifelse(treated ==1, 1, 0)) %>% 
  select(-c("healthshock")) %>% 
  select(DUID,FAMIDYR,starts_with("healthshock"),EXPTOT) %>% 
  pivot_longer(cols=starts_with("healthshock")) %>% # summary()
  filter(value == 1) %>% # keep spending in each category only if it is equal to 1
  mutate(name=factor(name,levels=c("healthshock_none","healthshock_any","healthshock_breast","healthshock_colon","healthshock_lung", "healthshock_pros"),labels=c("No Health Shock","Any Cancer","Breast Cancer","Colon Cancer","Lung Cancer", "Prostate Cancer"))) %>% 
  ggplot(aes(x = name, y = log(EXPTOT), fill = name)) + 
    geom_boxplot() + theme_classic() + 
  labs(x="", y="", fill ="Own Health Shock") + 
  theme(axis.text.x=element_text(angle=-30))

fig4 <- cohort2 %>% ungroup() %>% mutate(healthshock_none = ifelse(treated==0,1,0),
                                 healthshock_any = ifelse(treated ==1, 1, 0)) %>% 
  select(-c("healthshock")) %>% 
  select(DUID,FAMIDYR,starts_with("healthshock"),EXPTOT) %>% 
  pivot_longer(cols=starts_with("healthshock")) %>% # summary()
  filter(value == 1) %>% # keep spending in each category only if it is equal to 1
  mutate(name=factor(name,levels=c("healthshock_none","healthshock_any","healthshock_breast","healthshock_colon","healthshock_lung", "healthshock_pros"),labels=c("No Health Shock","Any Cancer","Breast Cancer","Colon Cancer","Lung Cancer", "Prostate Cancer"))) %>% 
  ggplot(aes(x = name, y = log(EXPTOT), fill = name)) + 
    geom_boxplot() + theme_classic() + 
  labs(x="", y="", fill ="Spousal Health Shock") + 
  theme(axis.text.x=element_text(angle=-30))

# What do we make from these? 
ggsave(fig3, file=here("fig3.png"), width=6, height=6, units="in", dpi=300)
ggsave(fig4, file=here("fig4.png"), width=6, height=6, units="in", dpi=300)
```

## Package Citations

```{r, include=FALSE}
print("=============================Works Cited=============================")
loadedNamespaces() %>%
map(citation) %>%
print(style = "text") # Adds citations for each package to end of .rmd file

knitr::write_bib(file = 'packages.bib') # Constructs a citation file for all packages used in this lecture.

# DON'T FORGET TO CITE YOUR PACKAGES IN YOUR PAPERS/ASSIGNMENTS. 
```
