---
title: "Lecture 11 Code"
author: "Alex Hoagland"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r preliminaries}
################################################################################
# I like to include several additional notes in the header of my files here:
#
# Last modified: 8/17/2022
#
### PURPOSE:
  # Lecture 11 code and output file
# 
### NOTES:
  # - uses the Tidyverse package and Dplyr
################################################################################


### 0. Load the packages we will need for this file ####
library(tidyverse) # load the installed package for each new session of R
library(broom)
library(readxl) # Read in data
library(modelsummary) # For making regression tables
library(causaldata) # Useful toy data sets
library(here) # Helpful in working with directories and projects
library(AER) # this package has lots of applied metrics packages
library(foreign) # Helpful for reading in data from Stata or other code languages 
library(gghighlight) # For figures 
library(lubridate) # For figures
library(stargazer) # For tables

# TWFE Packages
library(fixest)
library(TwoWayFEWeights)
library(bacondecomp)
library(DIDmultiplegt)
# library(did)
library(did2s)
# library(etwfe)

set.seed(03262020) 
  # Setting the seed helps to make random number generators give us the same numbers across machines
################################################################################
```

## Preliminaries

Let's set up the same data we used in Lecture 9, on the effectiveness of vaccine lotteries on COVID-19 vaccination rates across the United States. The code below organizes the data and produces the same naive TWFE regression specification we discussed there.

```{r twfe-naive}
mydata <- read.csv(here("us_state_vaccinations.csv"))
mydata$date <- as.yearmon(mydata$date)

# Trim out some "states" that we don't need
`%!in%` <- Negate(`%in%`)
mydata <- mydata %>% filter(location  %!in% c("American Samoa", "Bureau of Prisons", "Dept of Defense",
                                             "Federated States of Micronesia", "Guam", "Indian Health Svc",
                                             "Long Term Care", "Marshall Islands", "Northern Mariana Islands",
                                             "Puerto Rico", "Republic of Palau", "United States", "Veterans Health",
                                             "Virgin Islands"))

# Group to month level
mydata <- mydata %>% group_by(date, location) %>% 
  summarize(total_vaccinations_per_hundred = sum(total_vaccinations_per_hundred,na.rm=T),
            people_vaccinated_per_hundred = sum(people_vaccinated_per_hundred,na.rm=T),
            monthly_vaccinations_per_million = sum(daily_vaccinations_per_million,na.rm=T))

# Identify all treated states for our TWFE regression
regdata_dte <- mydata %>% mutate(treatdate = ifelse(location %in% c("Ohio", "New York State", "Maryland"), 2021.417,
                                                     ifelse(location %in% c("Massachusetts", "Michigan"), 2021.583,NA)), # Note the treatment date doesn't matter for non-treated states
                                 reltime = round((as.numeric(date) - treatdate)*12)) 
regdata_dte <- regdata_dte %>% mutate(state = ifelse(location %in% c("Ohio", "New York State", "Maryland", "Massachusetts", "Michigan"), 1, 0),
                                      logy = log(monthly_vaccinations_per_million+1))
  # Make sure to reassign "treated" states

dte <- feols(logy ~ i(reltime, ref = -1) | 
               date + location, data = regdata_dte)

# And use coefplot() for a graph of effects
iplot(dte,ref.line=-0.2, 
      ref.line.par=list(col="red",lty=1,lwd=2)) # How do we interpret this? 
```

## Diagnosing Negative Weights

We want to know to what extent negative weights are an issue in our data. To do this, there are two packages we can use. First, we can use `TwoWayFEWeights`:

```{r twfe-diagnostics-1}
regdata_dte <- regdata_dte %>% ungroup()%>% mutate(treated = ifelse(reltime >= 0,1,0))
twowayfeweights(df=regdata_dte, # Dataframe
                Y="logy", # Dep var
                G="location", # group identifier
                T="date", # Date var
                D="treated", # Whether state was treated
                cmd_type="feTR") # type of estimation to perform -- most commonly feTR
weighttable <- twowayfeweights(df=regdata_dte, # Dataframe
                               Y="logy", # Dep var
                               G="location", # group identifier
                               T="date", # Date var
                               D="treated", # Whether state was treated
                               cmd_type="feTR") # type of estimation to perform -- most commonly feTR
  # lots of negative weighting here!
```

### Goodman-Bacon Decomposition

The Goodman-Bacon decomposition can be implemented using `bacondecomp`:

```{r twfe-diagnostics-2}
regdata_dte <- regdata_dte %>% mutate(treated = ifelse(is.na(treated),0,treated))
regdata_dte <- regdata_dte %>% mutate(stateid = as.numeric(as.factor(location)),
                                      dateid=as.numeric(as.factor(date))) # Need a state id variable as a number, not string
bacon(formula = logy ~ treated, 
      data=regdata_dte,
      id_var="stateid",
      time_var="dateid",
      quietly=F)


```

## Alternative Estimators

### Stacked DID

To run stacked DID, we will create a stack for all cohorts that are treated at a certain point. 

``` {r stacked}
# first, identify all cohorts of treated observations
cohort_dates <- regdata_dte %>% filter(treated == 1) %>% 
  group_by(stateid) %>% 
  filter(date == min(date)) # %>% ungroup() %>% 
  # select(date) %>% unique()
# two cohorts: Jun and Aug 2021

# for each cohort, pull all treated at time t + those not treated between (t-4,t+4)
`%!in%` <- Negate(`%in%`)
stack_1 <- regdata_dte %>% 
  filter(dateid >= 2 & dateid <= 10) %>% # Appropriate timing
  filter(stateid %!in% c(22,23)) %>% # Clean controls
  group_by(stateid) %>% mutate(evertreated = max(treated)) %>% ungroup() %>% 
  mutate(reltime = ifelse(evertreated == 1, dateid - 6, NA))
  # Reset relative time for this cohort
  
stack_2 <- regdata_dte %>% 
  filter(dateid >= 4 & dateid <= 12) %>% 
  filter(stateid %!in% c(21,33,36)) %>% 
  group_by(stateid) %>% mutate(evertreated = max(treated)) %>% ungroup() %>% 
   mutate(reltime = ifelse(evertreated == 1, dateid - 8, NA))
  
# Reset relative time for this cohort

# create stacked data
stack <- rbind(stack_1,stack_2)

# run event study on this data
dte <- feols(monthly_vaccinations_per_million ~ i(reltime, ref = -1) | 
               dateid + stateid, data = stack)

# And use coefplot() for a graph of effects
iplot(dte,ref.line=-0.2, 
      ref.line.par=list(col="red",lty=1,lwd=2)) # How do we interpret this? 
```

### $DID_M$ Estimator

If we are willing to assume no dynamic treatment effects, we can correct for contamination using the estimator of de Chaisemartin and D'Haultfoeuille (2020; see also Imai and Kim, 2021). This estimator uses the `DIDmultiplegt` package:

```{r did-muliple-gt}
did_multiplegt(df=regdata_dte, Y="logy", G="stateid", T="date", D="treated")
# This gives us a point estimate of the treatment effect. What about inference? 

# Add boostraps! Obviously this takes time 
did_multiplegt(df=regdata_dte, Y="logy", G="stateid", T="date", D="treated", brep=100)

# Can also incorporate placebos to check this
didm_placebo <- did_multiplegt(df=regdata_dte, Y="logy", G="stateid", T="date", D="treated", 
                placebo=17) # Max number of placebos: T-2 --> more placebos = more time
placebos <- rep(NA, 17)
for (i in 1:17) {
  placebos[i] <- didm_placebo[2*i+2] # Pull placebo estimates
}
df <- data.frame(placebos[!is.na(placebos)]) # Remove all NANs, turn into data frame
df <- data.frame(t(df), "Placebos")
names(df) <- c("Placebo", "test")
ggplot(df,aes(x=Placebo))+geom_histogram(bins=5,color='darkslategray4',fill='darkslategray3')+
  geom_vline(xintercept=as.numeric(didm_placebo[1]),size=2,color='red')+
  theme_classic()+labs(x="Estimated Effect Size")
```

### Callaway and Sant'Anna (2021): The `did` package

Let's implement the CS estimator using the `did` package:

```{r csdid}
# Need a variable for first period of treatment (see below)
# Note: there has to be a cleaner way to make this work, but I haven't found it yet. 
regdata_dte <- regdata_dte %>%
  mutate(firsttreat = ifelse(treated==1,dateid,NA)) %>%
  group_by(stateid) %>% 
  mutate(firsttreat = min(firsttreat,na.rm=T)) %>% # first period where observation is treated
  mutate(firsttreat = ifelse(is.infinite(firsttreat),0,firsttreat)) # want untreated units to have a 0

csdid <- att_gt(yname='logy',
       tname='dateid',
       idname='stateid',
       gname='firsttreat', 
              # Note: needs to be a variable containing first period that each unit was treated 
              # (0 for untreated units)
       xformla = ~ total_vaccinations_per_hundred, # if there are covariates to include
       data=regdata_dte,
       control_group="notyettreated", # Default here is "nevertreated" only
       anticipation=2) # Can set # of periods anticipation might occur
summary(csdid) # how many of these are significant? 
df <- data.frame(csdid$group,csdid$t,csdid$att)

# Visualizations
ggplot(df,aes(csdid.att)) + 
  geom_histogram(fill='darkslategray3',color='darkslategray4',bins=15) +
  theme_classic()+labs(x="Estimated ATT",y="",title="ATT: All (g,t) cells")+
  geom_vline(xintercept = 0,size=1.5,color='red')
  # Look at all that heterogeneity!

# Heterogeneity across groups
ggplot(df,aes(csdid.att,group=factor(csdid.group),fill=factor(csdid.group))) + 
  geom_histogram(bins=15) +
  theme_classic()+labs(x="Estimated ATT",y="",title="ATT By Groups",fill="Group")+
  geom_vline(xintercept = 0,size=1.5,color='red')
  # Treated groups have different ATT distributions

# Heterogeneity across periods 
# Let's bin time periods
df <- df %>% mutate(bintime = ifelse(csdid.t <= 6, 1, ifelse(csdid.t <= 12, 2, 3)))
ggplot(df,aes(csdid.att,group=factor(bintime),fill=factor(bintime))) + 
  geom_histogram(bins=15) +
  theme_classic()+labs(x="Estimated ATT",y="",title="ATT By Periods",fill="Periods (binned)")+
  geom_vline(xintercept = 0,size=1.5,color='red')
  # Treated groups have different ATT distributions
```

Why are there only two groups here? Because CS stacks the treated groups into cohorts, and we only have two effective treatment dates! This tells us a lot about the usefulness of our event study so far.

Overall, what might we conclude about the effectiveness of these lotteries?

### Borusyak et al.(2021) and Gardner (2021)

Let's implement an alternative estimator using the `did2s` package -- technically this implements the Gardner (2021) estimation method, which is typically equivalent to that of Borusyak et al. (2021). This package is much faster and easier to use, however.

```{r did2s}
regdata_dte <- regdata_dte %>% mutate(reltime = ifelse(is.na(reltime),Inf,reltime))
  # Works better if you replace NAs with Infs for untreated group

regdata_dte <- regdata_dte %>% mutate(reltime = factor(reltime,
                                                       levels=c(-7,-6,-5,-4,-3,-2,-1,0,1,2,3,4,5,6,7,8,9,10,Inf)))
  # Need reltime to be a factor before being passed through feols().
  # Just note that the levels are out of order here

es <- did2s(data=regdata_dte,
            yname = "logy", first_stage = ~ total_vaccinations_per_hundred | stateid + dateid, 
            second_stage = ~i(reltime,ref=c(-1,Inf)), treatment = "treated", 
            cluster_var = "dateid")

fixest::iplot(es, main = "Event study: Staggered treatment", xlab = "Relative time to treatment", col = "steelblue", ref.line = -0.5)

# Compare to the original event study plot (dte) 
dte <- feols(logy ~ i(reltime, ref = c(-1, Inf)) | 
               date + location, data = regdata_dte)

fixest::iplot(list(es, dte), sep = 0.2, ref.line = -0.5,
      col = c("steelblue", "#82b446"), pt.pch = c(20, 18), 
      xlab = "Relative time to treatment", 
      main = "Event study: Staggered treatment (comparison)")


# Legend
legend(x=12, y=-.45, col = c("steelblue", "#82b446"), pch = c(20, 18), 
       legend = c("Two-stage estimate", "TWFE"))
```

## All Alternatives at Once

If a referee **really** insists on it, there is a way you can produce a graph with multiple TWFE coefficients altogether. This code is adapted from [this guide](https://kylebutts.com/did2s/reference/event_study.html), and produces a graph for the 5 most common DiD estimators. 

```{r combine-es}
# Note: this comes from the did2s package again 
out = event_study(
  data = regdata_dte,
  yname = "logy", 
  idname = "stateid",
  tname = "dateid",
  gname = "firsttreat", # Unit specific date of initial treatment
  estimator = "all" # Can ask for some of c("all", "TWFE", "did2s", "did", "impute", "sunab", "staggered")
)

plot_event_study(out)
```

## Extended TWFE Using Pooled OLS
Finally, we can implement Wooldridge (2021)'s approach for a pooled OLS. This doesn't have a package, but it doesn't need one! Remember that in principle, this estimator simply suggests that to avoid some of the negative aspects of the traditional TWFE-DID estimator, one should simply use a fully interaction/saturated set of dummies allowing for treatment effect heterogeneity by cohort and timing.

```{r pols}

# To run this, we just need interactions between all cohorts and periods. 
# We already have periods, let's add in cohorts
regdata_dte <- regdata_dte %>% mutate(cohortid = as.factor(firsttreat))
pols <- lm(logy ~ factor(cohortid) + factor(dateid) + factor(cohortid):factor(dateid), 
           data=regdata_dte)
summary(pols)
```

And that's it! The only trick is now to separate out the coefficients back into relative time and aggregate, or impose restrictions, in order to make this more interpretable. I'll leave these as an exercise. 

## Package Citations

```{r, include=FALSE}
print("=============================Works Cited=============================")
loadedNamespaces() %>%
map(citation) %>%
print(style = "text") # Adds citations for each package to end of .rmd file

knitr::write_bib(file = 'packages.bib') # Constructs a citation file for all packages used in this lecture.

# DON'T FORGET TO CITE YOUR PACKAGES IN YOUR PAPERS/ASSIGNMENTS. 
```
