---
title: "Lecture 9 Code"
author: "Alex Hoagland"
date: "November 16, 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r preliminaries}
### 0. Load the packages we will need for this file ####
library(tidyverse) # load the installed package for each new session of R
library(broom)
library(readxl) # Read in data
library(modelsummary) # For making regression tables
library(causaldata) # Useful toy data sets
library(here) # Helpful in working with directories and projects
library(AER) # this package has lots of applied metrics packages
library(foreign) # Helpful for reading in data from Stata or other code languages 
library(devtools)
library(marginaleffects) # To calculate marginal effects
library(pscl) # For hurdle models
library(nnet) # For multinomial logit models
library(knitr) # Alternative table package
library(kableExtra) # Alternative table package
library(fixest)
library(gtsummary) # Alternative table package

set.seed(03262020) 
  # Setting the seed helps to make random number generators give us the same numbers across machines
################################################################################
```

## Linear Probability Models

Suppose we are estimated in the effect of how recently someone donated blood on whether or not they donated in March 2007. *What's the question here?* Let's first estimate a linear probability model:

```{r lpm}
mydata <- FFTrees::blood # data on blood donation 
mydata <- mydata %>% rename(total_donated = total) 
mydata %>% summary() 

# Main (binary) outcome: did they donate in March 2007
m1 <- lm(donation.crit ~ recency + frequency, data=mydata)
  # how to interpret the recency dummy?
msummary(list("LPM"=m1),
         vcov=c("robust"),
         stars=c('*' = .1, '**' = .05, '***' = .01))

# Effect sizes in context? 
  # Note: each month of recency is associated with a 1 percentage point decline in the likelihood of donating. How relevant is that? 
mydata %>% select(donation.crit) %>% summarize(mean = mean(donation.crit)) %>% 
  mutate(percentage_change = -.01/mean * 100) # What does this mean in context?
```

One thing that we need to be concerned about with an LPM is predicted probabilities that fall outside of the unit interval (e.g., are not probabilities). Does that happen here? How does the model do for prediction generally?

```{r pred-prob}

# Look at predictions
mydata$pred_donation <- predict(m1, mydata)
hist(mydata$pred_donation) # what problems do you see here? 
summary(mydata) # First problem: some predictions are outside of the unit interval!
mydata %>% filter(pred_donation > 1.1) # What does it mean to have a 110% probability of donating?

hist(mydata$pred_donation) # Second problem: how do you go from a predicted 57% probability of Y=1 to predicting Y? 
mydata <- mydata %>% mutate(yhat = ifelse(pred_donation >= 0.5, 1, 0)) # What happens if we assume a cutoff (Yhat >= 50% means Yhat = 1)? How good is our match then?
table(mydata$donation.crit, mydata$yhat) # We incorrectly match 22.6% of the data! And we only predict a few donations. Can we play around with this (lower thresholds?)
```

## Logit and Probit Models

Instead of an LPM, let's try estimating logit and probit models of the same equation:

```{r logit}

# Logit
library(fixest) 
m2 <- feglm(donation.crit ~ recency + frequency, data=mydata,
          family='binomial'(link='logit')) # since we have binary data

# Probit
m3 <- feglm(donation.crit ~ recency + frequency, data=mydata,
          family='binomial'(link='probit')) # since we have binary data

msummary(list("LPM"=m1,"Logit"=m2,"Probit"=m3),
         vcov=c(rep("robust",3)),
         stars=c('*' = .1, '**' = .05, '***' = .01))
```

We can visualize the output of a logit/probit regression (somewhat; see below) using an odds plot

```{r odds-plot}
# Fitted regression "lines"
library(ggdist) # Cool extensions for ggplot
ggplot(mydata,aes(x=recency,y=donation.crit))+
  geom_dots(aes(side=ifelse(donation.crit == 1, "bottom", "top")),
            pch=19, color="gray70", scale=0.2) + 
  geom_smooth(method="glm", method.args = list(family = binomial(link = "logit")),
              size=1, color='green', se = FALSE) + 
  labs(x="Months since last donation", y="Pr(Donates)")

# Another option: coefplot (useful in many settings! )
library(fixest) # note: do not use the coefplot library! Much harder to use!
fixest::coefplot(m1,drop = "Constant") + theme_classic() # coefplot of LPM. Why do we like this?

# Now, let's do coefplot of logit
fixest::coefplot(m2,drop = "Constant") + theme_classic()
# See below for coefplot once converted to marginal effects
```

What do these coefficients mean? Why are the logit and probit so different?

### Non-Collapsibility and Odds Ratios
Remember that interpreting **odds ratios** is difficult due to the problem of non-collapsibility. To see this, consider a simple simulated example: 

```{r simulate-or}
n <- 1000000
x <- rbinom(n, 1, 0.5)
z <- rnorm(n, 0, 1) # note that x and z are uncorrelated by assumption
# True model: logit(P(Y=1)) = -2 + 1*X + 0.5*Z
linpred <- -2 + 1*x + 0.5*z
p <- 1 / (1 + exp(-linpred))
y <- rbinom(n, 1, p)
simdata <- data.frame(y, x, z)

simdata <- simdata %>% mutate(y1 = ifelse(row_number() <= 500000, y, NA), 
                              x1 = ifelse(row_number() <= 500000, x, NA), 
                              z1 = ifelse(row_number() <= 500000, z, NA))

# Model 1: unadjusted
sim_1 <- feglm(y1 ~ x1, data=simdata,
          family='binomial'(link='logit')) # since we have binary data

# Model 2: adjutsed for Z 
sim_2 <- feglm(y1 ~ x1 + z1, data=simdata,
          family='binomial'(link='logit')) # since we have binary data

# Model 3: add underlying data 
sim_3 <- feglm(y ~ x + z, data=simdata,
          family='binomial'(link='logit')) # since we have binary data

msummary(list("Unadjsuted"=sim_1,"Adjusted"=sim_2,"Adjusted, More Data"=sim_3),
         vcov=c(rep("robust",3)),
         stars=c('*' = .1, '**' = .05, '***' = .01))

# are the coefficients on x statistically different across sim_1 and sim_2?
b1 <- coef(sim_1)["x1"]
b2 <- coef(sim_2)["x1"]
se1 <- vcov(sim_1)["x1", "x1"]
se2 <- vcov(sim_2)["x1", "x1"]

# Approximate z-test for equality of coefficients (assuming independence)
z <- (b1 - b2) / sqrt(se1 + se2)
pval <- 2 * pnorm(-abs(z))
pval # coefficient has totally changed!

# So then the odds ratios change as well? 
b3 <- coef(sim_3)["x"]
or1 <- exp(b1)
or2 <- exp(b2)
or3 <- exp(b3)
c(or1,or2,or3)

# Do the marginal effects in each model change? 
me1 <- avg_comparisons(sim_1)$estimate
me2 <- avg_comparisons(sim_2)$estimate[1]
me3 <- avg_comparisons(sim_3)$estimate[1]
c(me1,me2,me3)
```

### Marginal Effects
Recall that we need to compare the **marginal effects** across regressions, as the actual coefficient magnitudes are meaningless without making the appropriate transformations. Let's calculate each type of marginal effect for the logit model (doing so for the probit model is very similar)

```{r marginal-fx}

# 1. Each individual marginal effect
allmarginaleffects <- comparisons(m2) # Gives you individual ME for all variables/observations
allmarginaleffects <- allmarginaleffects %>% 
  mutate(bin = ifelse(p.value > 0.1, 0, 
                      ifelse(p.value <= 0.1 & p.value > 0.05, 1, 
                        ifelse(p.value <= 0.05 & p.value > 0.01, 2, 3))))
allmarginaleffects %>% filter(term == "recency") %>% 
  ggplot(aes(x=estimate)) + 
  geom_histogram(alpha=0.4, fill='blue', color='blue') + 
  theme_classic() + 
  labs(x="Marginal Effect", y = "Value", fill = "Significance")

# 2. and 4. MER and MEM: use the datagrid() function
avg_comparisons(m2, newdata=datagrid(recency = 7, frequency = 4, time = 31))

# 3. AME (most common)
avg_comparisons(m2) # This gives you AME
# plot_slopes(marginaleffects(m2)) # Convenience plot 
```

So then what kind of model should we prefer? Let's report the LPM, probit, and logit together in a table, using the average marginal effect for each 

```{r model-comparison}
# average marginal effects
# for lm(), avg_slopes() returns the same substantive quantity as the coefficient
ame1 <- avg_slopes(m1, variables = c("recency", "frequency"), vcov = "HC1")
ame2 <- avg_slopes(m2, variables = c("recency", "frequency"), vcov = "HC1")
ame3 <- avg_slopes(m3, variables = c("recency", "frequency"), vcov = "HC1")

# table
modelsummary(
  list(
    "LPM"    = ame1,
    "Logit"  = ame2,
    "Probit" = ame3
  ),
  estimate  = "{estimate}{stars}",
  statistic = "({std.error})",
  stars = c("*" = .1, "**" = .05, "***" = .01),
  coef_map = c(
    "recency"   = "Recency",
    "frequency" = "Frequency"
  ),
  gof_omit = ".*"
) # which model should we prefer here? 
```

## Poisson Regression

Now, suppose that our outcome of interest is not whether or not an individual purchased insurance, but the number of health visits they selected in a year.

### Preliminaries

First, I will simulate some data and bake in some correlations. You can mostly ignore this code

```{r sim-data}
mydata$numvisits <- rpois(1340,2) # Base numvisits
# Now bake in some correlation
mydata <- mydata %>% 
  mutate(numvisits = ifelse(age >= 65,round(numvisits*1.5),numvisits)) %>%
  mutate(numvisits = ifelse(age < 35,round(numvisits*0.5),numvisits)) %>%
  mutate(numvisits = ifelse(sex == "male",round(numvisits*0.8),numvisits)) %>%
  mutate(numvisits = numvisits * bmi/31) %>%
  mutate(numvisits = ifelse(numvisits > 10, 10, numvisits)) %>% 
  mutate(numvisits = ifelse(smoker == 1, numvisits * 2, numvisits)) %>% 
  mutate(numvisits = numvisits * expenses/9000) %>% 
  mutate(numvisits = ifelse(numvisits > 10, 10, numvisits)) %>% 
  mutate(numvisits = ifelse(coverage == 1,round(numvisits*1.2),numvisits))
mydata$numvisits <- as.integer(mydata$numvisits)
hist(mydata$numvisits)
```

You can see that after the simulations, the distribution of the number of visits looks pretty similar to how it would look if we could use claims data.

### Poisson Regression

Let's use all of our variables to model the decision to choose health visits. In particular, suppose that we are after the impact of insurance coverage on total utilization (measured in visit counts)?

```{r poisson}
m_poisson <- glm(numvisits ~ coverage + age + sex + bmi + children + smoker,data=mydata,
                 family = "poisson"(link = "log")) # Why is our link function the log function?
m_naive <- lm(numvisits ~ coverage + age + sex + bmi + children + smoker,data=mydata) # Naive OLS
msummary(list("Naive"=m_naive,"Poisson"=m_poisson),
         vcov=c(rep("robust",2)),
         stars=c('*' = .1, '**' = .05, '***' = .01))
```

We interpret these regressions, again, by looking at the marginal effects (in this case, we can also look at rate ratios; the preference really depends on the field/journal more than the research question).

```{r me-p}
exp(m_poisson$coefficients)-1 # What is the marginal effect for coverage? What is the rate ratio?
```

### Testing for Dispersion

We also need to test for overdispersion in order to run a Poisson model. That is, we want to test whether the mean and variance of our sample data are equivalent.

```{r dispersion}
var(mydata$numvisits)/mean(mydata$numvisits) # Lots of overdisperson! Could bootstrap this to get a CI. 
```
