---
title: "Lecture 8 (Day 2) Code: Negative-Control Tests for an IV"
author: "Alex Hoagland"
date: "November 9, 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r header}
################################################################################
# Last modified: 2026 (Day 2 add-on to the Lecture 7 IV code)
#
### PURPOSE:
  # Day 2 code: negative-control ("falsification") tests to bolster an IV design
#
### NOTES:
  # - Picks up the SAME simulated blood-type / waiting-time data from Day 1.
  # - A negative control is a check where the honest answer should be "no effect."
  #   If the instrument moves something it causally cannot, exclusion/exogeneity
  #   is in doubt.
################################################################################


### 0. Load the packages we will need for this file ####
library(tidyverse)
library(broom)
library(modelsummary) # For making regression tables
library(AER)          # ivreg()

set.seed(03262020) # SAME seed as Day 1 so the data are identical
################################################################################
```

## Picking up from Day 1

These falsification tests run on exactly the same simulated data we built in the
Day 1 code (blood type as an instrument for transplant waiting time, with `age`
as the unseen confounder). The chunk below reproduces that data set so this file
knits on its own -- it is the Day 1 data-generating code, unchanged.

```{r reproduce-day1-data}
mydata <- data.frame(
  id = 1:10000,
  age = sample(seq(0,100,1),size=10000,replace=T),
  blood_type = sample(c("O_pos", "A_pos", "B_pos", "O_neg", "A_neg", "AB_pos", "B_neg", "AB_neg"), size=10000, replace=T,
                      prob=c(0.38, 0.34, 0.09, 0.07, 0.06, 0.03, 0.02, 0.01)),
  wait_time = rexp(n=10000) * 100
)
mydata <- mydata %>% mutate(wait_time = ifelse(blood_type == "A_pos",wait_time * 0.12,
                                        ifelse(blood_type == "A_neg", wait_time * 0.12,
                                        ifelse(blood_type == "B_pos", wait_time * 0.46,
                                        ifelse(blood_type == "B_neg", wait_time * 0.46,
                                        ifelse(blood_type == "AB_pos", wait_time * 0.05,
                                        ifelse(blood_type == "AB_neg", wait_time * 0.05,
                                        ifelse(blood_type == "O_pos", wait_time * 0.54,
                                               wait_time * 0.54))))))))
mydata <- mydata %>% mutate(mortality = ifelse(wait_time > 180, sample(c(0,1), size=10000, replace = T, prob = c(0.75, 0.25)),
                                        ifelse(wait_time > 90, sample(c(0,1), size=10000, replace = T, prob = c(0.8, 0.2)),
                                        ifelse(wait_time > 60, sample(c(0,1), size=10000, replace = T, prob = c(0.9, 0.1)),
                                               sample(c(0,1), size=10000, replace = T, prob = c(0.97, 0.03))))))
mydata <- mydata %>% mutate(wait_time = wait_time * abs(age-30)/sd(age),
                            mortality = round(mortality * abs(age-30)))
mydata[which(mydata$mortality>1),]$mortality <- 1

# The instrument, exactly as on Day 1: "rare" blood types (O/B) wait longer
mydata <- mydata %>% mutate(rare = ifelse(blood_type %in% c("O_pos","O_neg","B_pos","B_neg"), 1, 0),
                            wt_months = wait_time / 30)
```

## Negative-control outcome

The instrument (rare blood type) should affect **mortality** only *through* longer
waiting times. So if we invent a **placebo outcome** that has nothing to do with
waiting time, the reduced form of that placebo on the instrument should be a
precise **null**.

Here `placebo` is pure noise -- an event unrelated to the transplant pathway (think
of a cause of death, or an administrative event, that waiting time cannot touch).

```{r nc-outcome}
mydata <- mydata %>% mutate(placebo = rbinom(n(), 1, 0.10)) # independent of everything

# Reduced form on the REAL outcome: instrument SHOULD predict mortality
rf_real    <- lm(mortality ~ rare, data = mydata)
# Reduced form on the PLACEBO outcome: instrument should NOT predict it
rf_placebo <- lm(placebo   ~ rare, data = mydata)

msummary(list("Real outcome (mortality)" = rf_real,
              "Placebo outcome"          = rf_placebo),
         vcov = c("robust","robust"),
         stars = c('*' = .1, '**' = .05, '***' = .01))
# We WANT: a significant coefficient on the left, and a null on the right.
```

A significant coefficient in the placebo column would be a red flag: it would mean
the instrument is correlated with the outcome through some channel *other* than
waiting time -- i.e., the exclusion restriction is in trouble.

## Negative-control balance (predetermined covariate)

A credible instrument is as-good-as-random with respect to characteristics that
are fixed *before* treatment. Age is our confounder, and by construction blood type
was assigned independently of it -- so a balance regression of `age` on the
instrument should come back **null**.

```{r nc-balance}
bal <- lm(age ~ rare, data = mydata)
msummary(list("Balance: age ~ instrument" = bal),
         vcov = c("robust"),
         stars = c('*' = .1, '**' = .05, '***' = .01))
# A null here is reassuring: the instrument is not just picking up the confounder.
```

## What a *failing* negative control looks like

To see the diagnostic value, build a deliberately **bad** instrument that is
correlated with age (the confounder). It fails the balance test -- exactly the
signal you never want to see for your real instrument.

```{r nc-failing}
mydata <- mydata %>% mutate(bad_z = as.numeric(age > 50) )  # correlated with the confounder
bad_bal <- lm(age ~ bad_z, data = mydata)
msummary(list("Balance: age ~ BAD instrument" = bad_bal),
         vcov = c("robust"),
         stars = c('*' = .1, '**' = .05, '***' = .01))
# A large, significant coefficient => this "instrument" is entangled with the
# confounder and would violate exogeneity.
```

## Takeaway

Passing these checks does not *prove* the instrument is valid, but it removes some
of the most common ways it could be invalid. Failing localizes the problem (which
assumption, which confounder) and points to fixes: add controls, redefine the
instrument, or restrict the sample.
